State of the Practice for Medical Imaging Software Based on Open Source Repositories

### 2024-11-09 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Edited up to, but not including, Section 7

We present the state of the practice for the development of Medical Imaging
(MI) software based on data available in open source repositories. We selected
29 projects from 48 candidates and assessed 9 software qualities
(installability, correctness/ verifiability, reliability, robustness,
usability, maintainability, reusability, understandability, and
visibility/transparency) by answering 108 questions for each. Using the
Analytic Hierarchy Process (AHP) on the quantitative data, we ranked the MI
software.  The top five are \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji},
\textit{OHIF Viewer}, and \textit{ParaView}.  This is consistent with the
community's view, with four of these also appearing in the top five using
GitHub metrics (stars-per-year).  The quality and quantity of documentation
present in a project correlates quite well with its popularity.  Generally, MI
software is in a healthy state: in the repositories
we observed 88\% of the documentation artifacts recommended by research
software development guidelines, 100\% of MI projects use version control
tools, and developers appear to use the common quasi-agile research software
development process. However, the current state of the practice deviates from
existing guidelines as some recommended artifacts are rarely present
(like a test plan, requirements specification, code of conduct, and code style
guidelines), low usage of continuous integration (17\% of the projects), low
use of unit testing (about 50\% of projects), and room for improvement with
documentation. From developper interviews , we identified 7 concerns: lack
of development time, lack of funding, technology hurdles, correctness,
usability, maintainability, and reproducibility. We recommend:
increase effort on documentation,
increase testing by enriching datasets, increase continuous integration usage,
move to web applications, employ linters, use peer reviews, and design for
change.

@@

As well as state of the practice (SOP) for MI software, we would like to
understand the impact of the often cited gap between recommended
software engineering practices and that used for most research 
software~\cite{Storer2017}. Although scientists spend
a substantial proportion of their working times on software development
\cite{Hannay2009, Prabhu2011}, few are formally trained~\cite{Hannay2009}.

Our investigation is based on the following research questions:
\begin{enumerate}
\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
{open source} software projects exist? (Section~\ref{SecWhatProjects})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
quantitative measurements of each project's development practices, which
projects follow best practices? (Section~\ref{SecMeasurementResults})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
similar are the top projects identified in \rqref{RQ_HighestQuality}
to the most popular projects as viewed by the community?
(Section~\ref{Sec_VsCommunityRanking})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
do MI projects compare to general research software with respect to the
artifacts (documents, scripts and code) present in their repositories?
(Section~\ref{Sec_CompareArtifacts})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:] How
do MI projects compare to research software in general with respect to the use
of tools (Section~\ref{Sec_CompareTools}) for development
(\rqref{RQ_CompareToolsProjMngmnt}.a); and, project management
(\rqref{RQ_CompareToolsProjMngmnt}.b)?
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
How do MI projects compare to research software in general with respect to
principles, processes, and methodologies used?
(Section~\ref{Sec_CompareMethodologies})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
the pain points for developers working on MI software projects?
(Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
do the pain points of developers from MI compare to the pain points
for research software in general? (Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
developers what specific best practices are taken to address the pain points
(Section~\ref{painpoints})
and software quality concerns? (Section~\ref{sec:sq})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
What research software development practice could potentially address the
pain point concerns identified in \rqref{RQ_PainPoints}?
(Section~\ref{ch_recommendations})
\end{enumerate}

@@

Quality is defined as a measure of the excellence or worth of an entity.  As is
common practice, we do not think of quality as a single measure, but rather as
a set of measures.  That is, quality is a collection of different qualities,
often called ``ilities.''  For this study we selected 9 qualities to measure:
installability, correctness/ verifiability, reliability, robustness, usability,
maintainability, reusability, understandability, and visibility/transparency.
With the exception of installability, all the qualities are defined in Ghezzi
et al. (2003) \cite{GhezziEtAl2003}. Installability is defined as the effort
required for the installation and/or uninstallation of software in a specified
environment \cite{ISO/IEC25010}.

@@

The repository-based measurements are incomplete because they don't generally
capture the development process, the developer pain points, the perceived
threats to software quality, and the developers' strategies to address these
threats.  Therefore, part of our methodology involves interviewing developers.
We based our interviews on a list of 20 questions, which can be found in
[citation redacted for double blind].
%Smith et al. \cite{SmithEtAl2021}. 
Some questions are about the background of the software, the development teams,
the interviewees, and how they organize their projects.  We also ask about the
developer's understanding of the users. Some questions focus on the current and
past difficulties, and the solutions the team has found, or plan to try. We also
discuss documentation, both with respect to how it is currently done, and how it
is perceived. A few questions are about specific software qualities, such as
maintainability, understandability, usability, and reproducibility. The
interviews are semi-structured based on the question list.

@@

The top four software products \textit{3D Slicer}, \textit{ImageJ},
\textit{Fiji}, and \textit{OHIF Viewer} have higher scores in most criteria.
\textit{3D Slicer} has a score in the top two for all qualities; \textit{ImageJ}
ranks near the top for all qualities, except for correctness \& verifiability.
\textit{OHIF Viewer} and \textit{Fiji} have similar overall scores, with
\textit{Fiji} doing better in installability and \textit{OHIF Viewer} doing
better in correctness \& verifiability.  Given the installation problems, we may
have underestimated the scores on reliability and robustness for \textit{DICOM
Viewer}, but we compared it equally for the other seven qualities.

@@

\item [Correctness \& Verifiability] The packages with higher scores for
correctness and verifiability used a wider array of techniques to improve
correctness, and had better documentation to witness this.  For instance, we
looked for evidence of unit testing, and found evidence for only about half of
the projects. We identified five projects using continuous integration tools:
\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
\textit{OHIF Viewer}. The only requirements-related document we found was a road
map of \textit{3D Slicer}, which contained design requirements for upcoming
changes.

@@

Our ranking and GitHub popularity, at least for the top five projects, seems to
line up fairly well. However, we ranked some popular packages fairly low, such
as \textit{dwv}. This is because we were unable to build it locally, even though
we followed its installation instructions. However, we were able to use its web
version for the rest of the measurements. Additionally, this version did not
detect a broken DICOM file and instead displayed a blank image
(Section~\ref{SecMeasurementResults}).  \textit{DICOM Viewer} ranked low as we
were unable to install the NextCloud platform.

We weighted all qualities equally, which is not the likely (implicit) weight
that users implicitly use. To properly assess this would require a broad user
study.  Furthermore our measures of popularity are only \emph{proxies} which
are biased towards past rather than current preferences~\cite{Szulik2017}, as
these are monotonically increasing quantities. Finally there are often more
factors than just quality that influence the popularity of ``consumer''
products.

@@

We show the recommended artifacts in Table~\ref{Tbl_Guidelines}, one per row.
For a given row, a checkmark in one of the columns means that the corresponding
guideline recommends this artifact.  The last column shows whether the artifact
appears in the measured set of MI software, either not at all (blank), commonly
(C), uncommonly (U) or rarely (R).  We did our best to interpret the meaning of
each artifact consistently between guidelines and specific MI software, but the
terminology and the contents of artifacts are not standardized.  The challenge
even exists for the ubiquitous README file.  The content of README files shows
significant variation between projects~\cite{PranaEtAl2018}.  Some content is
reasonably consistent, with 97\% of README files contain at least one section
describing the `What' of the repository and 89\% offering some `How' content,
other categories are more variable.  For instance, information on
`Contribution', `Why', and `Who', appear in 28\%, 26\% and 53\% of the analyzed
files, respectively~\cite{PranaEtAl2018}.

@@

Table~\ref{artifactspresent} presents our measurements for MI software.
The table groups the artifacts by frequency
into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages).

Note that ``popularity'' in Table~\ref{Tbl_Guidelines} does not imply that these
oft recommended artifacts are the most important. Guidelines are often brief,
to encourage adoption, and thus even guidelines that mention the need for
installation instructions rarely mention uninstallation instructions.
Two items in Table~\ref{artifactspresent} do not appear in any guidelines:
\emph{Troubleshooting guide} and \emph{Developer's manual}.  However the
information within these documents overlaps with
the recommended artifacts.  Troubleshooting information often can be found in
a User Manual, while the information in a ``Developer's Manual'' is often
scattered amongst many other documents.

@@

This is unfortunate as when scientific developers are surveyed on their pain
points, Wiese et al.~\cite{WieseEtAl2019} found that software requirements and
management is the software engineering discipline that most hurts them,
accounting for 23\% of the technical problems reported by study participants.
Further adding to the misfortune, there is a widespread perception that up-front
requirements are impossible for research software \cite{CarverEtAl2007,
SegalAndMorris2008}. Fortunately a more agile approach to requirements is
feasible~\cite{Smith2016}, and research-software specific templates
exist~\cite{SmithEtAl2007}.

@@

Developers use software tools for user support, version control (VCS),
Continuous Integration and Deployment (CI/CD), documentation and project
management.  We summarize the tool usage in each of these categories, and
compare this to the usage by the research software community.

@@

\paragraph{Version control} Most teams interviewed (eight of nine) use GitHub
for version control, which makes it unsurprising that they also use pull
requests for managing community contributions. The experience of using GitHub
was reported as generally positive; some teams had actively migrated to GitHub
from other VCS.  This represents a considerable uptake since its rare use in
2006~\cite{Wilson2006}, which matches the uptake in other communities: 10 years
ago Nguyen et al~\cite{Nguyen-HoanEtAl2010} estimated that only 50\% of
research software projects used version control, and by 2018 this had jumped to
over 80\%~\cite{AlNoamanyAndBorghi2018} with many communities~\cite{Smith2018}
reaching 100\% already in 2018. In 2022, version control use sits at over
95\%~\cite{CarverEtAl2022}.

\paragraph{CI/CD} 
Only 17\% of MI projects use CI/CD tools. We estimated this using artifacts
present in the repository -- actual usage may be higher as some tools are
entirely external. This was the case for a study of Lattice Boltzmann Method
(LBM) software [citation redacted for blind review].  %\cite{Michalski2021}.
This contrasts with the high frequency of guidelines recommending continuous
integration
\cite{BrettEtAl2021, Brown2015, ThielEtAl2020, Zadka2018, vanGompelEtAl2016}.
A recent survey~\cite{CarverEtAl2022} of practioners (rather than projects)
suggests higher use of CI/CD with 54\% of respondents indicating that they use
it.

@@

\paragraph{Project management} Two types of tools were mentionned:
\begin{inparaenum}[i)]
\item trackers, including GitHub, issue trackers, bug trackers and Jira; and,
\item documentation tools, including GitHub, Wiki page, Google Doc, and
Confluence.
\end{inparaenum}
Of the named tools, interviewees mentioned GitHub 3 times, and each
of the other tools once each.

Based on~\cite{JungEtAl2022}, tool use for MI
software is similar to tool use for ocean modelling software.
Both use tools for editing, compiling, code management, testing, building, and
project management, but differ
in the latter's use of Kanban boards for project management.

>>>>

We present the state of the practice for the development of Medical Imaging (MI)
software based on data available in open source repositories. We selected 29
projects from 48 candidates and assessed 9 software qualities (installability,
correctness/verifiability, reliability, robustness, usability, maintainability,
reusability, understandability, and visibility/transparency) by answering 108
questions for each. Using the Analytic Hierarchy Process (AHP) on the
quantitative data, we ranked the MI software.  The top five are \textit{3D
Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{OHIF Viewer}, and
\textit{ParaView}.  This is consistent with the community's view, with four of
these also appearing in the top five using GitHub metrics (stars-per-year).  The
quality and quantity of documentation present in a project correlates quite well
with its popularity.  Generally, MI software is in a healthy state: in the
repositories we observed 88\% of the documentation artifacts recommended by
research software development guidelines, 100\% of MI projects use version
control tools, and developers appear to use the common quasi-agile research
software development process. However, the current state of the practice
deviates from existing guidelines as some recommended artifacts are rarely
present (like a test plan, requirements specification, code of conduct, and code
style guidelines), low usage of continuous integration (17\% of the projects),
low use of unit testing (about 50\% of projects), and room for improvement with
documentation. From developer interviews, we identified 7 concerns: lack of
development time, lack of funding, technology hurdles, correctness, usability,
maintainability, and reproducibility. We recommend: increase effort on
documentation, increase testing by enriching datasets, increase continuous
integration usage, move to web applications, employ linters, use peer reviews,
and design for change.

@@

As well as state of the practice for MI software, we would like to understand
the impact of the often cited gap between recommended software engineering
practices and the practices used for most research software~\cite{Storer2017}. Although
scientists spend a substantial proportion of their working times on software
development \cite{Hannay2009, Prabhu2011}, few are formally
trained~\cite{Hannay2009}.

Our investigation is based on ten research questions:
\begin{enumerate}
\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
{open source} software projects exist? (Section~\ref{SecWhatProjects})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
quantitative measurements of each project's development practices, which
projects follow best practices? (Section~\ref{SecMeasurementResults})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
similar are the top projects identified in \rqref{RQ_HighestQuality}
to the most popular projects as viewed by the community?
(Section~\ref{Sec_VsCommunityRanking})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
do MI projects compare to general research software with respect to the
artifacts (documents, scripts and code) present in their repositories?
(Section~\ref{Sec_CompareArtifacts})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:] How
do MI projects compare to research software in general with respect to the use
of tools (Section~\ref{Sec_CompareTools}) for development
(\rqref{RQ_CompareToolsProjMngmnt}.a); and, project management
(\rqref{RQ_CompareToolsProjMngmnt}.b)?
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
How do MI projects compare to research software in general with respect to
principles, processes, and methodologies used?
(Section~\ref{Sec_CompareMethodologies})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
the pain points for developers working on MI software projects?
(Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
do the pain points of developers from MI compare to the pain points
for research software in general? (Section~\ref{painpoints})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
developers what specific best practices are taken to address the pain points
(Section~\ref{painpoints})
and software quality concerns? (Section~\ref{sec:sq})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
What research software development practice could potentially address the
pain point concerns identified in \rqref{RQ_PainPoints}?
(Section~\ref{ch_recommendations})
\end{enumerate}

@@

Quality is defined as a measure of the excellence or worth of an entity.  As is
common practice, we do not think of quality as a single measure, but rather as
a set of measures.  That is, quality is a collection of different qualities,
often called ``ilities.''  For this study we selected 9 qualities to measure:
installability, correctness/verifiability, reliability, robustness, usability,
maintainability, reusability, understandability, and visibility/transparency.
With the exception of installability, all the qualities are defined in Ghezzi
et al. (2003) \cite{GhezziEtAl2003}. Installability is defined as the effort
required for the installation and/or uninstallation of software in a specified
environment \cite{ISO/IEC25010}.

@@

The repository-based measurements are incomplete because they don't generally
capture the development process, the developer pain points, the perceived
threats to software quality, and the developers' strategies to address these
threats.  Therefore, part of our methodology involves interviewing developers.
We based our interviews on a list of 20 questions, which can be found in
[citation redacted for double blind].
%Smith et al. \cite{SmithEtAl2021}. 
Some questions are about the background of the software, the development teams,
the interviewees, and how they organize their projects.  We also ask about the
developer's understanding of the users. Additional questions focus on the current and
past difficulties, and the solutions the team has found, or plan to try. We also
discuss documentation, both with respect to how it is currently done, and how it
is perceived. A few questions are about specific software qualities, such as
maintainability, understandability, usability, and reproducibility. The
interviews are semi-structured based on the question list.

@@

The top four software products \textit{3D Slicer}, \textit{ImageJ},
\textit{Fiji}, and \textit{OHIF Viewer} have higher scores in most criteria.
\textit{3D Slicer} has a score in the top two for all qualities; \textit{ImageJ}
ranks near the top for all qualities, except for correctness \& verifiability.
\textit{OHIF Viewer} and \textit{Fiji} have similar overall scores, with
\textit{Fiji} doing better in installability and \textit{OHIF Viewer} doing
better in correctness \& verifiability.  Given the installation problems, we may
have underestimated the scores on reliability and robustness for \textit{DICOM
Viewer}, but we compared it equally for the other qualities.

@@

\item [Correctness \& Verifiability] The packages with higher scores for
correctness and verifiability used a wider array of techniques to improve
correctness, and had better documentation to support this. For instance, we
looked for evidence of unit testing, and found evidence for only about half of
the projects. We identified five projects using continuous integration tools:
\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
\textit{OHIF Viewer}. The only requirements-related document we found was a road
map of \textit{3D Slicer}, which contained design requirements for upcoming
changes.

@@

Our ranking and GitHub popularity, at least for the top five projects, seems to
line up well. However, we ranked some popular packages fairly low, such
as \textit{dwv}. This is because we were unable to build it locally, even though
we followed its installation instructions. However, we were able to use its web
version for the rest of the measurements. Additionally, this version did not
detect a broken DICOM file and instead displayed a blank image
(Section~\ref{SecMeasurementResults}).  \textit{DICOM Viewer} ranked low as we
were unable to install the NextCloud platform.

We weighted all qualities equally, which is not the likely the same weighting
that users implicitly use. To properly assess this would require a broad user
study.  Furthermore our measures of popularity (like stars) are only \emph{proxies} which
are biased towards past rather than current preferences~\cite{Szulik2017}, as
these are monotonically increasing quantities. Finally there are often more
factors than just quality that influence the popularity of 
products.

@@

We show the recommended artifacts in Table~\ref{Tbl_Guidelines}, one per row.
For a given row, a checkmark in one of the columns means that the corresponding
guideline recommends this artifact.  The last column shows whether the artifact
appears in the measured set of MI software, either not at all (blank), commonly
(C) (20 to 29 ($>$67\%) packages), uncommonly (U) (10 to 19 (33-67\%) packages)
or rarely (R) (1 to 9 ($<$33\%) packages).  We did our best to interpret the
meaning of each artifact consistently between guidelines and specific MI
software, but the terminology and the contents of artifacts are not
standardized.  The challenge even exists for the ubiquitous README file.  The
content of README files shows significant variation between
projects~\cite{PranaEtAl2018}.  Some content is reasonably consistent, with 97\%
of README files contain at least one section describing the `What' of the
repository and 89\% offering some `How' content, other categories are more
variable.  For instance, information on `Contribution', `Why', and `Who', appear
in 28\%, 26\% and 53\% of the analyzed files, respectively~\cite{PranaEtAl2018}.

@@

Popularity in Table~\ref{Tbl_Guidelines} does not imply that these oft
recommended artifacts are the most important. Guidelines are often brief, to
encourage adoption, and thus even guidelines that mention the need for
installation instructions rarely mention uninstallation instructions.  Two items
in Table~\ref{artifactspresent}, which presents our measurements for the MI
software, do not appear in any guidelines: \emph{Troubleshooting guide} and
\emph{Developer's manual}.  However the information within these documents
overlaps with the recommended artifacts.  Troubleshooting information often can
be found in a User Manual, while the information in a ``Developer's Manual'' is
often scattered amongst many other documents.

@@

This is unfortunate as when scientific developers are surveyed on their pain
points, Wiese et al.~\cite{WieseEtAl2019} found that software requirements and
management is the software engineering discipline that most hurts them,
accounting for 23\% of the technical problems reported by study participants.
Further adding to the misfortune, there is a widespread perception that up-front
requirements are impossible for research software \cite{CarverEtAl2007,
SegalAndMorris2008}. Fortunately an iterative approach to requirements is
feasible~\cite{Smith2016}, and research-software specific templates
exist~\cite{SmithEtAl2007}.

@@

Developers use software tools for user support, version control,
Continuous Integration and Deployment (CI/CD), documentation and project
management.  We summarize the tool usage in each of these categories, and
compare this to the usage by the research software community.

@@

\paragraph{Version control} Most teams interviewed (eight of nine) use GitHub
for version control, which makes it unsurprising that they also use pull
requests for managing community contributions. The experience of using GitHub
was reported as generally positive; some teams had actively migrated to GitHub
from other version control systems.  This represents a considerable uptake since
its rare use in 2006~\cite{Wilson2006}, which matches the uptake in other
communities: 10 years ago Nguyen et al~\cite{Nguyen-HoanEtAl2010} estimated that
only 50\% of research software projects used version control, and by 2018 this
had jumped to over 80\%~\cite{AlNoamanyAndBorghi2018} with many
communities~\cite{Smith2018} reaching 100\% already in 2018. In 2022, version
control use sits at over 95\%~\cite{CarverEtAl2022}.

\paragraph{CI/CD} 
Only 17\% of MI projects use CI/CD tools. We estimated this using artifacts
present in the repository -- actual usage may be higher as some tools are
entirely external. This was the case for a study of Lattice Boltzmann Method
(LBM) software [citation redacted for blind review].  %\cite{Michalski2021}.
This contrasts with the high frequency of guidelines recommending continuous
integration
\cite{BrettEtAl2021, Brown2015, ThielEtAl2020, Zadka2018, vanGompelEtAl2016}.
A recent survey~\cite{CarverEtAl2022} of practitioners (rather than projects)
suggests higher use of CI/CD with 54\% of respondents indicating that they use
it.

@@

\paragraph{Project management} Two types of tools were mentioned:
\begin{inparaenum}[i)]
\item trackers, including GitHub, issue trackers, bug trackers and Jira; and,
\item documentation tools, including GitHub, Wiki page, Google Doc, and
Confluence.
\end{inparaenum}
Of the named tools, interviewees mentioned GitHub 3 times, and each
of the other tools once.

Tool use for MI software is similar to tool use for ocean modelling
software~\cite{JungEtAl2022}.  Both use tools for editing, compiling, code
management, testing, building, and project management, but differ in the
latter's use of Kanban boards for project management.
